Context-dependent consequences of including lagged effects in demographic models

### 2023-07-20 docs/ipm_comparison_ms.qmd
adding quarto version of ms

---
title: "Context-dependent consequences of including lagged effects in demographic models"
shorttitle: "Lagged Effects IPMs"
author:
  - Eric R. Scott:
      email: scottericr@gmail.com
      institute: [UF_WEC]
      correspondence: true
  - Emilio M. Bruna:

      email: embruna@ufl.edu
      institute: [UF_WEC, UF_LAS, BDFFP]
      correspondence: false
institute:
  - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - UF_LAS: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530  
  - BDFFP: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    # bookdown::html_document2:
    #   theme:
    #     version: 4
    #     bootswatch: journal
    #   number_sections: no
    #   fig_caption: yes
    #   pandoc_args:
    #   - --lua-filter=templates/scholarly-metadata.lua
    #   - --lua-filter=templates/author-info-blocks.lua
      # - -Fpandoc-crossref #required for cross-ref equations
    bookdown::pdf_document2:
      floatsintext: no
      linenumbers: yes
      draft: yes
      mask: no
      figurelist: no
      tablelist: no
      footnotelist: no
      fontsize: 12pt
      toc: false
      number_sections: no
      fig_caption: yes
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
    # bookdown::word_document2:
    #   linenumbers: yes
    #   number_sections: no
    #   fig_caption: yes
    #   reference_docx: "templates/template.docx" # Insert path for the DOCX file
    #   pandoc_args:
    #   - --lua-filter=templates/scholarly-metadata.lua
    #   - --lua-filter=templates/author-info-blocks.lua
    #   - -Fpandoc-crossref #required for cross-ref equations
linkcolor: blue
bibliography: references.bib
csl: "templates/the-american-naturalist.csl" # Insert path for the bib-style
keywords: |
  integral projection models; environmental stochasticity; lagged effects
highlights: |
  These are the highlights

editor_options:
  markdown:
    wrap: sentence
    canonical: true

# header-includes:
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyfoot{}
  - \fancyhead[R]{p. \thepage}
  - \fancyhead[L]{Draft - `r format(Sys.time(), '%d %B %Y')`}
  - \usepackage{sectsty} \sectionfont{\centering}
  - \usepackage[utf8]{inputenc} # This loads package to convert unicode 
  - \DeclareUnicodeCharacter{394}{$\Delta$} # convert unicode \u0394 to Delta
  # - \usepackage{sectsty} \subsectionfont{\centering}
  # - \AtBeginDocument{\let\maketitle\relax}
 #  - \usepackage{float} #to keep figure in the location placed instead of end
 #  - \floatplacement{figure}{H}
 #  - \usepackage{booktabs}
 #  - \usepackage{setspace} # line spacing
 #  - \raggedbottom
 # # - \usepackage{endfloat} #[nomarkers] excludes the {insert figure x around here] from main text. The others exclude the list of tables and figures. https://cs.brown.edu/about/system/managed/latex/doc/endfloat.pdf
 #  - \usepackage{tabu}
 #  - \usepackage{makecell}
 #  - \usepackage{pdflscape}
 #  - \usepackage{longtable}
 #  - \newcommand{\blandscape}{\begin{landscape}}
 #  - \newcommand{\elandscape}{\end{landscape}}
 #  - \setlength\parindent{22pt}
---

@@

Keywords: `r rmarkdown::metadata$keywords`

Running title: `r rmarkdown::metadata$shorttitle`

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	dev = "png",
	dpi = 300,
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>"
)
library(targets)
library(tidyverse)
library(here)
```

@@

Abstract text.

@@

# Introduction

@@

1.  Demographic models, such as IPMs and matrix models, are widely used for all kinds of stuff.

2.  It has long been recognized that there is the potential for lagged effects.
    Including lagged effects in models has been a major technical challenge, but there are now multiple approaches for doing so.
    The studies assessing the potential for lagged effects on vital rates find that they indeed appear to be prevalent [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].

3.  Lagged effects could be different in different habitat types (and this could be a big reason for differences between habitats)

# Methods

## Demographic Data

@@

## IPMs

@@

-   Add info about additional datasets (Emilio)
-->
```
In preliminary investigation we found that the survival and growth of plants was better explained by treating seedlings and mature plants separately.
Seedlings are physiologically different from small plants because they necessarily lack the underground reserves (of carbohydrates and meristems) that a small, mature plant may have.
Therefore, we used general IPMs to model population dynamics with seedlings treated as a separate discreet class not structured by size.
General IPMs allow for combinations of continuous and discrete states and transitions between them [@ellnerDatadrivenModellingStructured2016].

@@

# Results

For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (Table \@ref(tab:aic)).
<!--# No longer true if random effect = different smooth for each year -->

@@

The time to iterate the DLNM models is much higher than than deterministic and kernel-resampled.
The greater use of computational resources is likely a result of `predict()` being much slower for GAMs with 2D smooths because the number of knots is much higher compared to the GAMs used for the vital rates models in the determinsitic and kernel-resampled IPMs.
<!--# also note that this can't really be paralellized because each iteration depends on the previous one.  There's *maybe* some potential speed gains in using multiple cores with predict.bam(), but unlikely to be substantial. -->

```{r}
meta_df <- 
  tar_meta(store = here("_targets")) %>% 
  select(name, size, bytes, time, seconds) %>% 
  filter(str_detect(name, "^ipm_\\w+_[cf]{2}$")) %>% 
  mutate(minutes = seconds / 60,
         hours = minutes / 60)

meta_df %>% 
  separate(name, into = c("trash", "IPM", "Habitat")) %>% 
  select(-trash) %>% 
  select(IPM, Habitat, minutes) %>% 
  group_by(IPM) %>% 
  summarize(mean_time_min = round(mean(minutes), 2))
```

@@

# Discussion

@@

# Acknowledgments

@@

# CRediT Statement

@@

# Data Availability Statement

@@

```{r lambdas, results='asis'}
lambda_table %>% 
  mutate(habitat = toupper(habitat)) %>% 
  mutate(across(c(est, lower, upper), ~format(round(.,4)))) %>% 
  mutate(lambda = glue::glue("{est} ({lower}, {upper})")) %>%
  mutate(ipm = str_replace_all(
    ipm,
    c(
      "det" = "Deterministic",
      "stoch" = "Stochastic, kernel-resampled",
      "DLNM" = "Stochastic, parameter-resampled"
    )
  )) %>% 
  select(IPM = ipm, Habitat = habitat, "$\\lambda$" = lambda) %>% 
pandoc.table(
  digits = 5,
  keep.trailing.zeros = TRUE,
  justify = c("left", "left", "right"),
  full_width = FALSE,
  caption = "(\\#tab:lambdas) Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
)

@@

# References

>>>>

---
title: "Context-dependent consequences of including lagged effects in demographic models"
shorttitle: "Lagged Effects IPMs"  
authors:
  - name: Eric R. Scott
    email: scottericr@gmail.com
    affil-id: 1
    corresponding: true
  - name: Emilio M. Bruna
    email: embruna@ufl.edu
    affil-id: 1,2,3
affiliations:
  - id: "1"
    name: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - id: "2"
    name: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530
  - id: "3"
    name: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: 'last-modified'
date-format: '([draft:] D MMMM YYYY)'
linkcolor: blue
bibliography: references.bib
csl: "templates/the-american-naturalist.csl" # Insert path for the bib-style
# 
# highlights: |
#   These are the highlights
format: 
  pdf:
    number-sections: false
    colorlinks: true
    keeptex: true
    # the following line is to render the aaffiliations in pdf, which is not default in quarto, see 
    # https://stackoverflow.com/questions/75040607/why-do-affiliations-not-show-up-anywhere-in-the-pdf-output-of-quarto
    template-partials:  
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\normalsize}
        \usepackage{lineno}\linenumbers
        \usepackage{sectsty}\sectionfont{\centering}
        \usepackage{sectsty}\subsectionfont{\centering}

# When ready to submit, these will change to double space and add a 
# paragraph indent
# \usepackage{setspace}\doublespacing
# \setlength\parindent{22pt}

fig-cap-location: bottom
tbl-cap-location: top
geometry:
  - top=1in
  - left=1in
  - bottom=1in
  - right=1in
keywords: 
  - integral projection models
  - environmental stochasticity
  - lagged effects
---

@@

<!-- Keywords: `r rmarkdown::metadata$keywords` -->

<!-- Running title: `r rmarkdown::metadata$shorttitle` -->

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	dev = "png",
	dpi = 300,
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>"
)
library(targets)
library(tidyverse)
library(kableExtra)
library(here)
```

@@

Text of 150 words max summarizing this amazing paper.

@@

## Introduction

@@

### **Paragraph 1**

1.  Structured demographic models, such as Integral Projection Models (IPMs) and matrix models, have emerged as invaluable tools in ecology, evolution, and conservation biology.

They provide insights into \_\_\_\_, the evolution of life-history strategies, and life-history stages and ecological processes at which conservation intervention is likely to have the greatest impact on population growth rate.

Their flexibility hasmade them increasingly applicable to species with differtent life histories.

however, there are still a number of challenges to address, namely \_\_\_\_ , \_\_\_\_, and \_\_\_\_.

### **Paragraph 2**

1.  It has long been recognized that there is the potential for lagged effects.
2.  Including lagged effects in models has been a major technical challenge, but there are now multiple approaches for doing so.
3.  The studies assessing the potential for lagged effects on vital rates find that they indeed appear to be prevalent [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].

### **Paragraph 3**

1.  Lagged effects could be different in different habitat types.
2.  This could be a big reason for differences between habitats.

### **Paragraph 4**

1.  Here we...
2.  We conducted these analyses with both deterministic and stochastic IPMs.

## Methods

### *Demographic methods and data*

@@

Overview of the heliconia project and data

<!-- Add info about additional datasets (Emilio) -->

### *Construction of Integral Projection Models*

@@

-->
```
In preliminary investigation we found that the survival and growth of plants was better explained by treating seedlings and mature plants separately.
Seedlings are physiologically different from small plants because they necessarily lack the underground reserves (of carbohydrates and meristems) that a small, mature plant may have.
Therefore, we used general IPMs to model population dynamics with seedlings treated as a separate discreet class not structured by size.
General IPMs allow for combinations of continuous and discrete states and transitions between them [@ellnerDatadrivenModellingStructured2016].

@@

### *Statistical analyses*

All about the stats.

## Results

For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (Table \@ref(tab:aic)).\
<!--# No longer true if random effect = different smooth for each year -->

@@

The time to iterate the DLNM models is much higher than than deterministic and kernel-resampled.

The greater use of computational resources is likely a result of `predict()` being much slower for GAMs with 2D smooths because the number of knots is much higher compared to the GAMs used for the vital rates models in the determinsitic and kernel-resampled IPMs.
<!--# also note that this can't really be paralellized because each iteration depends on the previous one.  There's *maybe* some potential speed gains in using multiple cores with predict.bam(), but unlikely to be substantial. -->

@@

## Discussion

@@

## Acknowledgments

@@

## CRediT Statement

@@

## Data Availability Statement

@@

```{r time, results='asis'}
meta_df <- 
  tar_meta(store = here("_targets")) %>% 
  select(name, size, bytes, time, seconds) %>% 
  filter(str_detect(name, "^ipm_\\w+_[cf]{2}$")) %>% 
  mutate(minutes = seconds / 60,
         hours = minutes / 60)

meta_df_tbl<-meta_df %>% 
  separate(name, into = c("trash", "IPM", "Habitat")) %>% 
  select(-trash) %>% 
  select(IPM, Habitat, minutes) %>% 
  group_by(IPM) %>% 
mutate(IPM = str_replace_all(
    IPM,
    c(
      "det" = "Deterministic",
      "stoch" = "Stochastic, kernel-resampled",
      "dlnm" = "Stochastic, parameter-resampled"
    ))) %>%
  summarize(mean_time_min = round(mean(minutes), 2)) %>%
  rename("IPM Type" = IPM) %>% 
  rename("mean time (min.)" = mean_time_min)

kbl(meta_df_tbl, 
    booktabs = T,
    align = "lc",
    caption = "table caption") %>% 
  kable_styling(full_width = F,
                position = "center")

```{=tex}
\newpage
\newpage
```
```{r lambdas, results='asis'}
lambda_table %>% 
  mutate(habitat = toupper(habitat)) %>% 
  mutate(across(c(est, lower, upper), ~format(round(.,4)))) %>% 
  mutate(lambda = glue::glue("{est} ({lower}, {upper})")) %>%
  mutate(ipm = str_replace_all(
    ipm,
    c(
      "det" = "Deterministic",
      "stoch" = "Stochastic, kernel-resampled",
      "DLNM" = "Stochastic, parameter-resampled"
    )
  )) %>% 
  select(IPM = ipm, Habitat = habitat, "$\\lambda$" = lambda) %>% 
pandoc.table(
  digits = 5,
  keep.trailing.zeros = TRUE,
  justify = c("left", "left", "right"),
  full_width = FALSE,
  caption = "(\\#tab:lambdas) Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
)

@@

## Literature Cited
